---
title: "Huhu_analysis"
author: "Mafalda S. Baptista"
date: "2023-10-27"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 10, fig.height = 6)
```


```{r Libraries and seed}

# Libraries
library("tidyverse")       # data wrangling and visualisation

library("knitr")           # markdown output control

library("phyloseq")        # analysis of microbial communities

library("vegan")           # multivariate analysis

library("patchwork")       # plots together

library("plotly")          # Interactive plots

library("here")            # set the path to the folder 

set.seed(57)
```

### BACTERIA
```{r Load the data}

# Load the data

huhu_bac <- readRDS("huhu_bac.rds")
huhu_bac
```


```{r Get metadata}

# Load the metadata
metadata_bac <-  read.csv("Huhu_bac_metadata.csv", row.names = 1, header = TRUE, stringsAsFactors = TRUE)

# Create a phyloseq object
huhu_bac <- phyloseq(otu_table(huhu_bac), 
                          tax_table(huhu_bac),
                          sample_data(metadata_bac))
huhu_bac
```


```{r Rename ASVs}

# Inspect the ASV data - it should be a sequence
otu_table(huhu_bac)[1:10, 1:10]


# We want to name the ASV to something shorter, but before lets store the seqs in phyloseq too

dna <- Biostrings::DNAStringSet(taxa_names(huhu_bac))
names(dna) <- taxa_names(huhu_bac)
huhu_bac <- merge_phyloseq(huhu_bac, dna)
taxa_names(huhu_bac) <- paste0("ASV_", seq(ntaxa(huhu_bac)))

# Re-inspect the data - it should be ASV_Number
otu_table(huhu_bac)[1:10, 1:10]

# And re-inspect the phyloseq - now the seqs are present
huhu_bac
```


#### Filtering by taxonomy
```{r Filter by taxonomy}

# Show available ranks in the dataset
rank_names(huhu_bac)

# How many domains? (Kingdom in the taxonomy table)
table(tax_table(huhu_bac)[, "Kingdom"], exclude = NULL) 

# How many Phyla?
table(tax_table(huhu_bac)[, "Phylum"], exclude = NULL) 

# Remove known erroneous assignments
huhu_bac_f <- subset_taxa (huhu_bac, (Order !="Chloroplast") | is.na(Order)) # (=559 ASVs)
huhu_bac_f <- subset_taxa (huhu_bac_f, (Family != "Mitochondria") | is.na(Family)) # (=469 ASVs)
huhu_bac_f

# Inspect Phyla by abundance

# Compute prevalence of each ASV
prevdf <- apply(X = otu_table(huhu_bac_f),
               MARGIN = ifelse(taxa_are_rows(huhu_bac_f), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this
prevdf <- data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(huhu_bac_f),
                    tax_table(huhu_bac_f))

# Plot prevalence
ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(huhu_bac_f), color = Phylum)) + 
  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +
  xlab("Total Abundance") + 
  ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```

Two Archaea. We will filter based on low abundance, and not on taxonomy.

### Sequences and ASVs data

```{r ASVs}

# Inspect number of reads by ASVs

colSums.asv.df <- data.frame(colSums(otu_table(huhu_bac_f))) %>%
  rownames_to_column("ASV")

ggplot(colSums.asv.df, aes(x = reorder(ASV, -colSums.otu_table.huhu_bac_f..), y = colSums.otu_table.huhu_bac_f..)) + 
  geom_bar(stat = "identity") +
  ylab("ASVs") +
  xlab("") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

colSums.asv.df %>% summarise(
    max = max(colSums.otu_table.huhu_bac_f..),
    median = median(colSums.otu_table.huhu_bac_f..),
    min = min(colSums.otu_table.huhu_bac_f..) 
    ) %>% 
      kable()
```
The number of ASVs reads varied between 2 (as expected) and 41112 for ASV_1. 


```{r Reads}
# Inspect number of reads by sample

rowSums.asv.df <- merge(data.frame(rowSums(otu_table(huhu_bac_f))), data.frame(sample_data(huhu_bac_f)), by = "row.names", all.x=TRUE) 

ggplot(rowSums.asv.df, aes(x = reorder(Row.names, -rowSums.otu_table.huhu_bac_f..), y =
               rowSums.otu_table.huhu_bac_f..)) +
  geom_col() + 
  coord_flip() +
  ylab("Reads") +
  xlab("")

rowSums.asv.df %>% summarise(
    max = max(rowSums.otu_table.huhu_bac_f..),
    median = median(rowSums.otu_table.huhu_bac_f..),
    min = min(rowSums.otu_table.huhu_bac_f..) 
    ) %>% 
      kable()

```

The number of reads in each Sample varied between 6974 and 33893, with a median value of 23255. If we remove samples with less than 10000 reads we would loose 20P_Bac, which clearly has less reads that all the others. But we will keep it until we check the rarefaction curves


```{r IonTorrent Runs}

ggplot(rowSums.asv.df, aes(x = reorder(Row.names, -rowSums.otu_table.huhu_bac_f..), y =
               rowSums.otu_table.huhu_bac_f..)) +
  geom_col(aes (fill = PlateNr)) + 
  coord_flip() +
  ylab("Reads") +
  xlab("")

``` 

We don't see any obvious difference between runs


```{r Interactive plot}

rowSums.asv.df <- rowSums.asv.df %>%
  arrange(desc(rowSums.otu_table.huhu_bac_f..))

rowSums.asv.df$Row.names

yform <- list(categoryorder ="array",
              categoryarray = c("TAP_Bac", "38P_Bac", "42P_Bac", "37P_Bac", "24P_Bac", "01C_Bac", "37C_Bac", "41P_Bac", "06P_Bac", "41C_Bac",
                                "01P_Bac", "20C_Bac", "42C_Bac", "06C_Bac", "24C_Bac", "38C_Bac", "TAC_Bac", "20P_Bac"))

plot_ly(rowSums.asv.df, x= ~rowSums.otu_table.huhu_bac_f.., y = ~Samples, 
        type = "bar", 
        color = ~PlateNr,
        text = ~rowSums.otu_table.huhu_bac_f..) %>%
  layout(yaxis = yform)

```


```{r Rarefaction}

# Plot rarefaction curves for each sample
# rarecurve(otu_table(huhu_bac_f), step = 100, 10000, lwd=2, ylab="ASVs")

# It is not working anymore, and it seems to be an issue w/ updating phyloseq
# https://stackoverflow.com/questions/74530039/error-vegan-rarecurve-error-in-asx-matrixi-j-drop-false-subscript

m_huhu_bac_f <- otu_table(huhu_bac_f)
class(m_huhu_bac_f) <- "matrix"

rarecurve(m_huhu_bac_f, step = 100, sample = 10000, lwd = 2, ylab = "ASVs")

```

Looks like 20P_Bac has reached a plateau. We will keep this sample. 


```{r Save}

# Save
saveRDS(huhu_bac_f, "huhu_bac_phyloseq.rds")
```

Save the phyloseq object for further analysis


**Revised paper**/
For the paper we ended up removing TAC_Bac and TAP_Bac because these samples were collected in Hillcrest and were deemed not comparable to the other samples, collected in Claudelands.

```{r Rarefaction revised}

huhu_bac_f_rev <- subset_samples(huhu_bac_f, Samples != "TAC_Bac" & Samples != "TAP_Bac")

# Plot rarefaction curves for each sample

m_huhu_bac_f_rev <- otu_table(huhu_bac_f_rev)
class(m_huhu_bac_f_rev) <- "matrix"

raremax <- min(rowSums(m_huhu_bac_f_rev))
rarecurve(m_huhu_bac_f_rev, step = 100, sample = raremax, lwd = 2, ylab = "ASVs", main = "A")


rowSums.asv.df <- merge(data.frame(rowSums(otu_table(huhu_bac_f_rev))), data.frame(sample_data(huhu_bac_f_rev)), by = "row.names", all.x=TRUE) 

ggplot(rowSums.asv.df, aes(x = reorder(Row.names, -rowSums.otu_table.huhu_bac_f_rev..), y =
               rowSums.otu_table.huhu_bac_f_rev..)) +
  geom_col() + 
  coord_flip() +
  ylab("Reads") +
  xlab("") +
  theme_bw()

rowSums.asv.df %>% summarise(
    max = max(rowSums.otu_table.huhu_bac_f_rev..),
    median = median(rowSums.otu_table.huhu_bac_f_rev..),
    min = min(rowSums.otu_table.huhu_bac_f_rev..) 
    ) %>% 
      kable()

```


### FUNGI
```{r Load the data f}

# Load the data

huhu_f <- readRDS("huhu_f.rds")
huhu_f
```


```{r Get metadata f}

# Load the metadata
metadata_f <-  read.csv("Huhu_f_metadata.csv", row.names = 1, header = TRUE, stringsAsFactors = TRUE)

# Create a phyloseq object
huhu_f <- phyloseq(otu_table(huhu_f), 
                          tax_table(huhu_f),
                          sample_data(metadata_f))
huhu_f
```


```{r Rename ASVs f}

# Inspect the ASV data - it should be a sequence
otu_table(huhu_f)[1:10, 1:10]


# We want to name the ASV to something shorter, but before lets store the seqs in phyloseq too

dna <- Biostrings::DNAStringSet(taxa_names(huhu_f))
names(dna) <- taxa_names(huhu_f)
huhu_f <- merge_phyloseq(huhu_f, dna)
taxa_names(huhu_f) <- paste0("ASV_", seq(ntaxa(huhu_f)))

# Re-inspect the data - it should be ASV_Number
otu_table(huhu_f)[1:10, 1:10]

# And re-inspect the phyloseq - now the seqs are present
huhu_f
```


#### Filtering by taxonomy
```{r Filter by taxonomy f}

# Show available ranks in the dataset
rank_names(huhu_f)

# How many domains? (Kingdom in the taxonomy table)
table(tax_table(huhu_f)[, "Kingdom"], exclude = NULL) 

# How many Phyla?
table(tax_table(huhu_f)[, "Phylum"], exclude = NULL) 


# Inspect Phyla tby abundance

# Compute prevalence of each ASV
prevdf <- apply(X = otu_table(huhu_f),
               MARGIN = ifelse(taxa_are_rows(huhu_f), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

# Add taxonomy and total read counts to this
prevdf <- data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(huhu_f),
                    tax_table(huhu_f))

# Plot prevalence
ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(huhu_f), color = Phylum)) + 
  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +
  xlab("Total Abundance") + 
  ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + theme(legend.position="none")
```

For now we will keep the NA Phyla

### Sequences and ASVs data

```{r ASVs f}

# Inspect number of reads by ASVs

colSums.asv.df <- data.frame(colSums(otu_table(huhu_f))) %>%
  rownames_to_column("ASV")

ggplot(colSums.asv.df, aes(x = reorder(ASV, -colSums.otu_table.huhu_f..), y = colSums.otu_table.huhu_f..)) + 
  geom_bar(stat = "identity") +
  ylab("ASVs") +
  xlab("") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

colSums.asv.df %>% summarise(
    max = max(colSums.otu_table.huhu_f..),
    median = median(colSums.otu_table.huhu_f..),
    min = min(colSums.otu_table.huhu_f..) 
    ) %>% 
      kable()
```
The number of ASVs reads varied between 2 (as expected) and 37245 for ASV_1. 


```{r Reads f}
# Inspect number of reads by sample

rowSums.asv.df <- merge(data.frame(rowSums(otu_table(huhu_f))), data.frame(sample_data(huhu_f)), by = "row.names", all.x=TRUE) 

ggplot(rowSums.asv.df, aes(x = reorder(Row.names, -rowSums.otu_table.huhu_f..), y =
               rowSums.otu_table.huhu_f..)) +
  geom_col() + 
  coord_flip() +
  ylab("Reads") +
  xlab("")

rowSums.asv.df %>% summarise(
    max = max(rowSums.otu_table.huhu_f..),
    median = median(rowSums.otu_table.huhu_f..),
    min = min(rowSums.otu_table.huhu_f..) 
    ) %>% 
      kable()

```

The number of reads in each Sample varied between 2105 and 33779, with a median value of 23255. We have a great discrepancy in the number of reads for sample 01C_F and 38C_F and all other samples.


```{r IonTorrent Runs f}

ggplot(rowSums.asv.df, aes(x = reorder(Row.names, -rowSums.otu_table.huhu_f..), y =
               rowSums.otu_table.huhu_f..)) +
  geom_col(aes (fill = PlateNr)) + 
  coord_flip() +
  ylab("Reads") +
  xlab("")

``` 

We don't see any obvious difference between runs

```{r Interactive plot f}

rowSums.asv.df <- rowSums.asv.df %>%
  arrange(desc(rowSums.otu_table.huhu_f..))

rowSums.asv.df$Row.names

yform <- list(categoryorder ="array",
              categoryarray = c("01C_F", "38C_F", "24P_F", "38P_F", "37C_F", "41P_F", "20P_F", "42P_F", "01P_F", "TAC_F", "06C_F", "06P_F",
                                "TAP_F", "37P_F", "42C_F", "41C_F", "24C_F", "20C_F"))

plot_ly(rowSums.asv.df, x= ~rowSums.otu_table.huhu_f.., y = ~Samples, 
        type = "bar", 
        color = ~PlateNr,
        text = ~rowSums.otu_table.huhu_f..) %>%
  layout(yaxis = yform)

```

```{r Rarefaction f}

# Plot rarefaction curves for each sample
#rarecurve(otu_table(huhu_f), step = 100, 10000, lwd=2, ylab="ASVs")

m_huhu_f <- otu_table(huhu_f)
class(m_huhu_f) <- "matrix"

rarecurve(m_huhu_f, step = 100, sample = 10000, lwd =2, ylab = "ASVs")

```

Looks like all the samples reached a plateau, so let's use this as a criteria. Sample TAC_F is not looking good, but we will remove it for the analysis. 

```{r Save f}

# Save
saveRDS(huhu_f, "huhu_f_phyloseq.rds")
```

Save the phyloseq object for further analysis

**Revised paper**/
For the paper we ended up removing TAC_Bac and TAP_Bac because these samples were collected in Hillcrest and were deemed not comparable to the other samples, collected in Claudelands.

```{r Rarefaction f revised}

huhu_f_rev <- subset_samples(huhu_f, Samples != "TAC_F" & Samples != "TAP_F")

# Plot rarefaction curves for each sample

m_huhu_f_rev <- otu_table(huhu_f_rev)
class(m_huhu_f_rev) <- "matrix"

raremaxf <- min(rowSums(m_huhu_f_rev))
rarecurve(m_huhu_f_rev, step = 100, sample = raremaxf, lwd = 2, ylab = "ASVs", main = "B")


rowSums.asv.df <- merge(data.frame(rowSums(otu_table(huhu_f_rev))), data.frame(sample_data(huhu_f_rev)), by = "row.names", all.x=TRUE) 

ggplot(rowSums.asv.df, aes(x = reorder(Row.names, -rowSums.otu_table.huhu_f_rev..), y =
               rowSums.otu_table.huhu_f_rev..)) +
  geom_col() + 
  coord_flip() +
  ylab("Reads") +
  xlab("") +
  theme_bw()

rowSums.asv.df %>% summarise(
    max = max(rowSums.otu_table.huhu_f_rev..),
    median = median(rowSums.otu_table.huhu_f_rev..),
    min = min(rowSums.otu_table.huhu_f_rev..) 
    ) %>% 
      kable()
```

Revised rarefaction plot for A) Bacteria and B) Fungi

```{r Rarefaction total plot }


par(mfrow = c(1,2))

rarecurve(m_huhu_bac_f_rev, step = 100, sample = raremax, lwd = 2, ylab = "ASVs", main = "A", label = FALSE)

rarecurve(m_huhu_f_rev, step = 100, sample = raremaxf, lwd = 2, ylab = "ASVs", main = "B", label = FALSE)

```
